---
title: "Homework5"
author: "Eric Song"
format: html
editor: visual
---

```{r,include=FALSE}
library(tidyverse)
```

## Task 1

**Original Code**

```{r}
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
str(d3)
print(nrow(d3)) # 382 students
```

**Using inner join().**   

```{r}
joined_d3 <- inner_join(d1,d2, 
                        by = c("school", "sex", "age", "address", "famsize",                           "Pstatus","Medu", "Fedu", "Mjob", "Fjob", "reason",                            "nursery", "internet")
                        )

str(joined_d3)
```

Many to many relationship warning on row 79.  Our data is also split between non joined variables.      



**All variables but no G1,G2,G3,Paid, and absences.**

```{r}
final_d1 <- inner_join(d1 |>
                         select(-G1, -G2, -G3, -paid, -absences), d2 |>
                            select(-G1, -G2, -G3, -paid, -absences))
str(final_d1)
```

**Picking School,Sex, Famsize, and Reason as our categorical variables.**
```{r}
final_d1 <- final_d1 %>%
  mutate(
    school = as.factor(school),
    sex = as.factor(sex),
    famsize = as.factor(famsize),
    reason = as.factor(reason)
  )
str(final_d1)
```

## Task 2

#### Categorical Variables

**One-way contingency table**
```{r}
one_way_tbl <- table(final_d1$school)
one_way_tbl
one_way_tbl[1]
```

There are 287 students going to GP.

**Two-way contingency table**
```{r}
two_way_tbl <- table(final_d1$school,final_d1$sex)
two_way_tbl
two_way_tbl[1,1]
```

153 Female students go to GP.  

**Three-way contingency table**
```{r}
three_way_tbl <- table(final_d1$school,final_d1$sex,final_d1$famsize)
three_way_tbl
three_way_tbl[1,1,1]
```

115 Female Students that go to GP come from a family greater than 3.  


**Filter method:**
```{r}
filter_d1 <- final_d1 %>% filter(school == "GP")
cond_two_way <-table(filter_d1$sex,filter_d1$famsize)
cond_two_way
```

**Three way table method (using previous three way variable):**
```{r}
cond_three_way <- three_way_tbl["GP",,]
cond_three_way
```

**Using group_by(),summarize(),and pivot_wider().**

```{r}
new_two_way <- final_d1 |>
                  drop_na(school,sex) |>
                  group_by(school,sex) |>
                  summarize(count=n()) |>
                  pivot_wider(names_from=sex,values_from=count)
new_two_way
```

**Stacked bar graph**
```{r}
ggplot(data= final_d1 |> drop_na(school,sex),aes(x = school, fill = sex)) +
  geom_bar() + 
  labs(x = "School",
    y = "# of Students"
  )
```
**Side by Side Graph.**
```{r}
ggplot(data= final_d1 |> drop_na(school,sex),aes(x = school, fill = sex)) +
  geom_bar(position = "dodge") + 
  labs(x = "School",
    y = "# of Students"
  )
```

#### Numeric Variables
